---
title: "The Ugly Truth about Covid-19 Situation in Indonesia"
description: |
  This article aims to do a fact checking of some officials' controversial statements related to Covid-19 situation in Indonesia.
author:
  - name: Dewi Lestari Amaliah
    url: {}
date: 08-31-2020
bibliography: references.bib
output:
  distill::distill_article:
    self_contained: false
categories:
  - COVID-19
  - Data Analysis
draft: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# Install the tidycovid19 package first if its have not been installed.
#devtools::install_github("joachim-gassen/tidycovid19")

library(tidyverse)
library(tidycovid19)
library(timevis)
library(DT)
library(kableExtra)
library(formattable)
library(plotly)
library(data.table)
library(gridExtra)
library(lubridate)
```


```{css, echo = FALSE}
.vis-item {
  background-color: #A3E4D7 ;
}


html, body, p {
  
  font-family: 'Lato', sans-serif;
  font-weight: 350;
  line-height: 1.3; 
  font-size: 15px;
  color: black;
  font-style: normal;
  
}

table.dataTable tbody td {
    padding: 12px 10px;
}


d-article table th, d-article table td {
    font-size: 13px;
    padding: 12px 10px;
}

.vis-item .vis-item-content {
    white-space: nowrap;
    box-sizing: border-box;
    padding: 5px;
    font-size: 13px;
}

```





## Introduction 

The statements from officials, experts, and leaders matter during an ongoing crisis, in this case, is Covid-19. It is because the messages that being told would determine people's expectation, fear, and reactions to the pandemic [@westman]. The problem is that not all the information regarding the crisis is true. The World Health Organization (WHO) even stated that the urgent measure to tackle the "coronavirus infodemic", the term for Covid-19 misinformation, should be taken immediately [@infodemic]. This infodemic would lead to public confusion about what information and which source to trust [@infodemic2].

One of the examples of the communication problem was shown by the U.S. government. While the U.S. Center for Disease Control and Prevention (CDC) stated that the situation would change quickly and the future is still unknown, President Trump claimed a convincing timeline of vaccine development [@westman]. 

Indonesia, the most populous country in South East Asia, also experiences a similar problem. Since the beginning of the emergence of Covid-19 in this country, the officials has made several controversial statements shown Figure \@ref(fig:statement-timeline). One of the reasons why these statements are considered controversial is because as per 28th August 2020, the country is reported has the highest death toll in the region, with 7.169 deaths, more than doubled the second place, Philippines, with 3.325 deaths [@csis]. In addition, it is feared that these positive statements will make the public less careful about the virus. Especially now that Indonesia has applied the "New Normal" policy that intended reopening of all the social and economic sectors’ followed with strict hygiene and health protocol [@newnormal]. 

Therefore, this article aims to do fact checking of these controversial claims using multiple data sources related to Covid-19. 


<center>
```{r statement-timeline, fig.cap= "The timeline of controversial statement made by the government of Indonesia"}

statement <- tribble(~id, ~content, ~start, ~end,
                1, "Covid-19 virus is weaker in Indonesia due to relatively high temperature","2020-04-03", NA, 
                2, "Covid-19 spread has gone down, so the new normal scheme could be applied", "2020-06-02", NA, 
                3, "The number of Covid-19 red zone has dropped", "2020-08-18", NA, 
                4, "There will not be a second wave of Covid-19 outbreak in Indonesia", "2020-08-13", NA, 
                5, "The Covid-19 positive rate has gone down and recovery rate has improved greatly to almost 66%", "2020-08-14", NA, 
                6, "Covid-19 in Indonesia is now undercontrol", "2020-08-19", NA, 
                )

timevis(statement, showZoom = FALSE, height = "350px", width = "800px")


```

</center>




## Data Description


```{r}

# download data for all countries available in the package and all of the compiled data
merged_data <- download_merged_data(cached = TRUE, silent = TRUE)

# filter the data only for Indonesia 
idn_data <- merged_data %>% filter(iso3c == "IDN") %>%
  select(-ecdc_cases, -ecdc_deaths, -region, -income, -timestamp) %>%
  filter(confirmed != 0)
```

```{r}
countries <- merged_data %>%
  group_by(country) %>%
  count()

ncountries <- nrow(countries)
```


This article is utilized the compiled dataset from `tidycovid19` R package [@tidycovi19]. This package comprises of Covid-19 related data from various source at the country-day level. Initially, there are `r ncountries` countries covered in this dataset, but for this article purpose, I used only Indonesia data start from the first case was recorded. Besides, to achieve the aim of this article, I mainly used the confirmed, fatality, and recovered cases data, which are compiled from John Hopkins University Center for System Science and Engineering (JHU CSSE) [@jhu]; the lockdown measure data from ACAPS [@acaps]; the testing data from Our World in Data [@testing]; and the population data from the World Bank [@population]. 


That being said, I also provided most of the variables in the compiled data in the package in Table \@ref(tab:idn-table). Note that I excluded the region variable (East Asia and Pacific), the income (upper middle income), and the time stamp variable (because it is the same with date variable) from the display.

<br>

```{r idn-table}

DT::datatable(idn_data, 
              caption = "Indonesia's Covid-19 related data", 
              options = list(
  pageLength=5, scrollX='400px'), filter = "top")

```


<details><summary> The definition of each variable can be seen here. </summary>

```{r}
vd <- tidycovid19_variable_definitions

kable(vd, col.names = c("Variable Name", "Source", "Definition"), caption = "Variable definition of Covid-19 dataset") %>% kable_styling(bootstrap_options = c("striped", "hover"))

```

</details>



## Analysis and Findings

The analysis is performed mainly using the functions in `tidyverse` [@tidyverse] package in R [@r]. I also used `DT` [@DT], `kableExtra` [@kableExtra], `timevis` [@timevis], `plotly` [@plotly], `gridExtra` [@gridExtra], and `formattable` [@formattable] packages. Since `tidycovid19` only provides cumulative cases, death, recovered, and tests, I wrangled the dataset to obtain the daily number. I also mutated new variables, namely infection rate, positive rate, and testing rate. The definition of those variables is as follows:

- Infection rate: According to @infectionrate, there are three ways of calculating the infection rate. The formula I used here is the number of infected people divided by the total population multiplied by the constant variable, which is 1,000,000 people, this figure is calculated in a daily level. 
- Positive rate: It is the fraction of tests that have a positive result. The calculation is the number of cumulative confirmed cases divided by the total tests (percentage) [@positiverate].
- Testing Rate : The number of test per 1,000,000 population at daily level. 

<br>

**Limitation**

Since the results of the Covid-19 test in Indonesia were not released on the same day, there should be a lag between the number of cases and the number of tests used to calculate the positive rate. However, the length of lag between the test time and the results varies between test locations, making it difficult to determine the right lag. Therefore, in the calculation, the positive rate is based on the cumulative number. 


```{r daily}
idn_data_new <- idn_data %>% select(iso3c, country, date, confirmed, deaths, recovered, total_tests, population, pop_density, lockdown) %>% 
  mutate(lag_confirmed = lag(confirmed, 1),
         lag_deaths = lag(deaths, 1),
         lag_recovered = lag(recovered, 1),
         lag_test = lag(total_tests, 1),
         daily_confirmed = case_when(date == "2020-03-02" ~ confirmed,
                                     date > "2020-03-02" ~ confirmed - lag_confirmed),
         daily_deaths = case_when(date == "2020-03-11" ~ deaths,
                                     date > "2020-03-11" ~ deaths - lag_deaths,
                                  date < "2020-03-11" ~ deaths),
         daily_recovered = case_when(date == "2020-03-10" ~ recovered,
                                     date > "2020-03-10" ~ recovered - lag_recovered,
                                     date < "2020-03-10" ~ recovered),
         daily_tests = case_when(date == "2020-03-16" ~ total_tests,
                                 date > "2020-03-16" ~ total_tests - lag_test),
         infection_rate = daily_confirmed / population * 1000000,
         positive_rate = confirmed / total_tests * 100,
         test_rate = daily_tests / population *1000000)

```


In addition, I would not check all of the statements in Figure \@ref(fig:statement-timeline). I would only check the statements that could be answered using this dataset one by one.

<br>

> **Statement 1: Covid-19 spread has gone down, so the new normal scheme could be applied.**

This claim was stated by the Coordinating Minister for Maritime Affairs and Investment, Luhut Binsar Pandjaitan, on June, 2nd 2020. He claimed that the decision to reopen the economy and ease the mobility restriction, which is called the "New Normal", is based on data that the spread of the virus has declined and the number of new cases has gone down to 600 cases per day on average [@detik].  

**The Fact**

There are two main points that I would check in this statement. First, is the declining of the average of new cases per day and the second is the declining of the virus spread. 


```{r}
June2 <- idn_data_new %>% filter(date == "2020-06-02")
```

The number of daily case shown in Figure \@ref(fig:ir-cases-plot) suggests that it was true that the number of cases as of June, 2 2020 was `r June2$daily_confirmed`, as claimed. Previous days, the number of cases was also in the range of 600s. However, we could not really say that the cases was declining as until this statement was made, June 2 2020, the trend was positive. 

Regarding the spreading of the virus, the Minister did not say its definition. Thus, I would assume that it is the infection rate, since the spread is supposed to be regarding the number of people who are likely to get infected by the virus. The infection rate shown in Figure \@ref(fig:ir-cases-plot) suggests that the spread of the virus was not declining. It was true that the infection rate had decreased at the end of May, compared to the third week of May, but again, the trend line shows that the spread of the virus was still increasing. 

The government, should have seen the bigger picture, instead of the spike shown in only a short time period. Hence, we could argue that this claim is busted. 


```{r data-for-plot}
plot_dat <- idn_data_new %>% filter(date <= "2020-06-02") 

plot_dat <- plot_dat %>% 
  mutate(index = seq(1: nrow(plot_dat)))

loess_ir <- loess(infection_rate ~ index, data = plot_dat)
loess_case <- loess(daily_confirmed ~ index, data = plot_dat)

smoothed_ir <- predict(loess_ir)
smoothed_case <- predict(loess_case)

plot_dat <- cbind(plot_dat, smoothed_ir, smoothed_case)

```

<center>

```{r ir-cases-plot, fig.cap = "Daily confirmed cases and infection rate of Covid-19 in Indonesia from 2020-03-2 until 2020-06-02"}

fig1 <- plot_ly(plot_dat, 
                x = ~date, 
                y = ~infection_rate, 
                name = 'Infection rate per 1,000,000 people', 
                type = 'scatter', 
                mode = 'lines', 
                line = list(color = '#F18BA2')) 
fig1 <- fig1 %>% add_trace(y = ~smoothed_ir, 
                           name = 'Trend of infection rate (loess method)', 
                           mode = 'lines',
                           line = list(color = "#AE5F7E"))
  

fig2 <- plot_ly(plot_dat, 
                x = ~date, 
                y = ~daily_confirmed, 
                name = 'Number of confirmed cases', 
                type = 'scatter',
                mode = 'lines',
                line = list(color = "#60B0DD")) 
fig2 <- fig2 %>% add_trace(y = ~smoothed_case, 
                           name = 'Trend of confirmed cases (loess method)',
                           mode = 'lines',
                           line = list(color = "#21618C")) 

fig <- subplot(fig2, fig1) %>%
  layout(legend = list(orientation = "h",
                       x = 0.05, y = -0.1))

fig

```

</center>



```{r}
# create new variable: the goverment policy, to see the change the before and after new normal
idn_data_new <- idn_data_new %>% mutate(govt_policy = case_when(date <= "2020-06-04" ~ "Before New Normal",
                                                                date > "2020-06-04" ~ "After New Normal"))
```


Despite this positive trend, Indonesian government has applied the "New Normal" scheme started from 5th June 2020 [@newnormalstarted]. Therefore, I was curious about how the statistics changed after this scheme put in.
I used median value as the summary statistics because the distribution of the indicators, which are the daily cases, daily fatality, daily recovered, daily testing, infection rate, and testing rate, are mostly skewed, especially before "New Normal" phase. According The Australian Bureau of Statistics (ABS) [@abs], the median is more robust to skewed data and is sometimes the favorable measure of central tendency when the distribution is asymmetric. 

<details><summary> Expand here to see the distribution.</summary>

```{r geom-density, fig.height = 15, fig.width = 9}
p1 <- ggplot(idn_data_new) +
  geom_density(aes(x = daily_confirmed, fill = govt_policy), alpha = 0.3) +
  facet_wrap(~govt_policy) +
  ggtitle("Daily confirmed cases")

p2 <- ggplot(idn_data_new) +
  geom_density(aes(x = daily_deaths, fill = govt_policy), alpha = 0.3) +
  facet_wrap(~govt_policy) +
  ggtitle("Daily fatality")

p3 <- ggplot(idn_data_new) +
  geom_density(aes(x = daily_recovered, fill = govt_policy), alpha = 0.3) +
  facet_wrap(~govt_policy) +
  ggtitle("Daily recovered cases")

p4 <- ggplot(idn_data_new) +
  geom_density(aes(x = daily_tests, fill = govt_policy), alpha = 0.3) +
  facet_wrap(~govt_policy) +
  ggtitle("Daily number of peope tested")

p5 <- ggplot(idn_data_new) +
  geom_density(aes(x = infection_rate, fill = govt_policy), alpha = 0.3) +
  facet_wrap(~govt_policy) +
  ggtitle("Daily infection rate")

p6 <- ggplot(idn_data_new) +
  geom_density(aes(x = test_rate, fill = govt_policy), alpha = 0.3) +
  facet_wrap(~govt_policy) +
  ggtitle("Daily testing rate")

p7 <- ggplot(idn_data_new) +
  geom_density(aes(x = positive_rate, fill = govt_policy), alpha = 0.3) +
  facet_wrap(~govt_policy) +
  ggtitle("Cumulative positive rate")


grid.arrange(p1, p2, p3, p4, p5, p6, p7, ncol= 1)


```


</details>



```{r indicator-median}


# create the summary table of median 
summary_table <- idn_data_new %>%
  group_by(govt_policy) %>%
  summarise(`Median of daily confirmed cases` = round(median(daily_confirmed, na.rm = TRUE), 3), 
            `Median of daily fatality` = round(median(daily_deaths, na.rm = TRUE), 3),
            `Median of daily recovered` = round(median(daily_recovered, na.rm = TRUE), 3),
            `Median of daily people tested` = round(median(daily_tests, na.rm = TRUE), 3),
            `Median of daily infection rate` = round(median(infection_rate, na.rm = TRUE), 3),
            `Median of daily testing rate` = round(median(test_rate, na.rm = TRUE), 3),
            `Median of cumulative positive rate` = round(median(positive_rate, na.rm = TRUE), 3)) 
 
summary_table_pivot_longer <- summary_table %>% pivot_longer(c(2:8), 
                                                             names_to = "Indicator", 
                                                             values_to = "Values")

summary_table_pivot_wider <- summary_table_pivot_longer %>% 
  pivot_wider(names_from = "govt_policy", 
              values_from = "Values")

# change the order of the column in rename the column

col_order <- c("Indicator", "Before New Normal", "After New Normal")

summary_table_new <-  summary_table_pivot_wider[, col_order]

```
<br>

All indicators in Table 2. show an increase after the "New Normal" policy was implemented. If we look at the number of confirmed cases and the infection rate, we will conclude that after the "New Normal", there has been a drastic increase in Covid-19 cases in Indonesia by about five times. Likewise with the death toll and recovery cases, the number was increasing because the number of cases was increasing.

In fact, this does not necessarily explain the true condition because the number of cases is greatly influenced by the number of people tested. If we look at the number of people tested, there was an increase after the New Normal period, even though the rate is only 43 people per 1,000,000 population per day. Hence, increasing cases makes sense.

Furthermore, if we look at a positive rate, we can see that looking at the number of positive cases only is not enough. In fact, there was a decrease in positive rates after the "New Normal". However, bear in mind that the positive rate here is cumulative figure. In addition, speaking of positive rate and "New Normal", the WHO recommended that the positive rate from a comprehensive testing program should be at or below 5% for at least 14 days to ease the restriction and resume the economic and social activity.[@positiverate] How about Indonesia? Figure \@ref(fig:positiverate-plot) shows the answer that Indonesia did not follow the WHO reccommendation because the positive rate was still above 10 percent when the "New Normal" began. 

<br>

Table 2. Covid-19 indicators before and after 'New Normal' policy in Indonesia

```{r sum-table, fig.cap= "Covid-19 indicators before and after 'New Normal' policy in Indonesia (Daily figure)"}

# edit the display of the table


formattable(summary_table_new, align =c("l","r","r"), list(
  `Indicator` = formatter("span",
                               style = ~ formattable::style(color = "black")), 
  `After New Normal`= formatter("span", style = ~ formattable::style(color = ifelse(`After New Normal` >`Before New Normal`, "#B14555", "steelblue")),
                    ~ icontext(ifelse(`After New Normal` >`Before New Normal`,"arrow-up", "arrow-down"), `After New Normal`))

)) 
  
```



<br>

> **Statement 2: The COVID-19 positive rate has gone down and recovery rate has improved greatly to almost 66%**

This contend was stated also by the Coordinating Minister for Maritime Affairs and Investment on August, 14th 2020. [@statement2]
Luhut Claimed that the handling of Covid-19, shown by the declining positive rate, has improved. Moreover, the recovery rate is nearly 66 percent. 


**The Fact**

```{r recovrate}

Aug14 <- idn_data_new %>% filter(date == "2020-08-14")

recov_delta1 <- round(Aug14$recovered/ Aug14$confirmed * 100, 2)

```

Figure \@ref(fig:positiverate-plot) shows that the positive rate has decreased when compared to the start of the pandemic in the country. However, in this figure, we can see that there is a slight inverted curve which indicates that the positive rate is increasing again.

Regarding 66 percent recovery claims, I tried to divide the recovered cases by the confirmed cases on that date and got a result of `r recov_delta1` percent. This means that the claim is true. 

But wait, is this good news? According to the health experts, using the recovery rate as an indicator to assess the performance of handling Covid-19 is not ideal. This is because first, the recovery rate is calculated by dividing the number of discharged patients with total Covid-19 patients. However, those who are hospitalized can either recover or die, so they cannot be included in the calculation. [@expert1] Secondly, 99% of people infected with COVID recovers, so the number of recovery rate is not really useful.[@expert2]


<center>

```{r positiverate-plot, fig.cap= "The positivity rate of Covid-19 in Indonesia"}
cuts1 <- data.frame(Ref="New Normal started", vals=c(ymd("2020-06-05")))
cuts2 <- data.frame(Ref="Statement 2 made", vals=c(ymd("2020-08-14")))
cuts3 <- data.frame(Ref="Statement 4 made", vals=c(ymd("2020-08-20")))
cuts <- rbind(cuts1, cuts2, cuts3)
  
fig3 <- ggplot(idn_data_new) +
  geom_line(aes(x = date, y = positive_rate), color = "#48C9B0") +
  geom_smooth(aes(x = date, y = positive_rate),
              method = "loess", se = FALSE, color = "#5C8F8D") +
  geom_vline(data = cuts , aes(xintercept=vals, color= Ref),
             show.legend = FALSE) +
  geom_text(mapping = aes(x = vals,
                          y = 0,
                          label = Ref,
                          hjust = 0,
                          vjust = -0.5,
                          angle = 90),
              data = cuts) +
  ylab("Postive rate (%)") +
  xlab("Date") +
  theme_light()

fig3
```
</center>



> **Statement 3: There will not be a second wave of COVID-19 outbreak in Indonesia**

The Coordinating Minister for Maritime Affairs and Investment did seem to like to preach "good news". On 13 August 2020, he claimed that Indonesia would not experience the second wave of Covid-19.[@secondwave]


**The Fact**

The fact of this claim is pretty straight forward. It is true. According to Dr.Anthony Fauci, the director of the U.S. National Institute of Allergy and Infectious Diseases, the second wave happens if the number of confirmed cases goes up and then comes back down to reach low single digits.[@secondwavedef] The example of this case is New Zealand, who had 24 days streak of zeros case before getting the new cases again. Hence, how could Indonesia experience the second wave, if the first wave has not yet reached its peak?  


<center>

```{r second-wave, fig.cap= "The number of confirmed cases daily in Indonesia"}


fig4 <- ggplot(idn_data_new) +
  geom_line(aes(x = date,
                y = daily_confirmed), color = "#60B0DD") +
  geom_smooth(aes(x = date,
                y = daily_confirmed),
              method = "loess", se = FALSE, color = "#21618C") +
  ylab("Daily confirmed cases") +
  xlab("Date") +
  theme_light()


ggplotly(fig4)


```

</center>


> **Statement 4: COVID-19 in Indonesia is now under control**

Now, you could guess who made the statement. Yes, it is correct. This statement was, again, told by the Minister for Maritime Affairs and Investment on 19 August 2020. 

**The Fact**

```{r}

postiverateaugdat <- idn_data_new %>% filter(date == "2020-08-19")

positiverate19aug <- postiverateaugdat$positive_rate
```


The fact is this claim is might be incorrect, at least according to the WHO criteria. According to the WHO in May, one of the indicator that the epidemic is under control is that positive rate is less than 5% [@postiveratewho]. As per date of this statement, the positive rate in Indonesia is `r positiverate19aug` percent.


## Conclusion

3 out of 4 of Luhut's claims, which were checked with data from various sources, are correct, in numerical terms. However, if we look at the bigger picture and the context, these claims are actually busted. For example, if the trend is seen in its entirety since the beginning of the pandemic, the cases and the spread of Covid-19 in Indonesia cannot be said as declined. Besides, contextually, the usage of recovered cases as an indicator to claim the success of Covid-19 handling can not be justified. 

The ugly truth about the Covid-19 situation in Indonesia is that the cases are still increasing, the testing rates per billion population are still small, and the positive rate is still above 5 percent. However, the uglier truth is that the "false positive claims" have been made by the person who should have known better.

