---
title: "COVID and Conflict: Why the pandemic could be worse in Afghanistan?"
description: |
  This article aims to convey the reason why the conflicted country, in this case is Afghanistan, might be hit harder by COVID-19.
author:
  - name: Dewi Lestari Amaliah
    url: {}
date: 09-15-2020
output:
  distill::distill_article:
    self_contained: false
    toc: true
bibliography: references2.bib
categories:
  - COVID-19
  - Data Analysis
draft: TRUE
preview: war.jpg
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(tidycovid19)
#library(plotly)
#library(hrbrthemes)
#library(waffle)
library(gt)
#library(DT)
#library(kableExtra)
library(gghighlight)
#library(crosstalk)
library(ggrepel)
library(scales)
```


```{css, echo = FALSE}

html, body, p {
  
  font-family: 'Lato', sans-serif;
  font-weight: 200;
  line-height: 1.3; 
  font-size: 1.0em;
  color: black;
  font-style: normal;
  
}

table.dataTable tbody td {
    padding: 12px 10px;
}


d-article table th, d-article table td {
    font-size: 13px;
    padding: 12px 10px;
}

```



## Introduction

Since being declared as a global pandemic by the World Health Organization (WHO) on 12 March 2020, COVID-19 has affected 213 countries, including the conflict-torn country, Afghanistan [@countries]. This country has its first confirmed case recorded on 24 February 2020 in Herat Povince after a citizen who had just returned from Iran contracted with the virus [@firstcase].

Viruses are indeed indiscriminate. Unfortunately, the presence of COVID-19 amid this war-shattered country seems to rub salt into the wound. According to the United Nations Office for the Coordination of Humanitarian Affairs (OCHA) data, as of 6 September 2020, there were already 151,190 people displaced by the conflict [@ocha]. Further, in 2019, civilian casualties due to the conflict has surpassed 100,000 [@vulnerable]. The Secretary-General’s Special Representative for Afghanistan and head of the UN Assistance Mission in Afghanistan (UNAMA) said as follows:  

> "Almost no civilian in Afghanistan has escaped being personally affected in some way by the ongoing violence" - *Tadamichi Yamamoto* [@vulnerable] 

In other words, the Afghan people are already vulnerable before COVID-19 attacks. 

This article aims to convey the reasons why the conflict brings the pandemic getting worse in Afghanistan. The analysis would be focused on two aspects, namely health care and economic vulnerability.  


## Data Source

In doing analysis, I used several data sets from various sources. For the COVID-19 situation in Afghanistan, I used the data of confirmed cases and deaths from John Hopkins University Center for System Science and Engineering (JHU CSSE) [@jhu] and the population data from the World Bank [@population]. I also used the mobility data from Google COVID-19 Community Mobile Report [@googlecmr] to analyze the population vulnerability. These data have been compiled in the `tidycovid19` R package by Joachim Gassen [@tidycovi19], hence I directly downloaded it from this package. For the health care vulnerability analysis, I mainly used the data of health sector profile from the Afghanistan National Statistics and Information Authority [@nsia] and the government national budget data from the Afghanistan Ministry of Finance [@mof].  


```{r covid-data}
# download data for all countries available in the package and all of the compiled data
merged_data <- download_merged_data(cached = TRUE, silent = TRUE)

# filter the data only for Afghanistan 
afghan_data <- merged_data %>% filter(iso3c == "AFG") %>%
  select(-ecdc_cases, -ecdc_deaths) %>%
  filter(confirmed != 0)
```

```{r afghan-data}
afghan_data_new <- afghan_data %>% select(iso3c, country, date, deaths, confirmed, population, gcmr_retail_recreation,
                                          gcmr_grocery_pharmacy, gcmr_parks, gcmr_transit_stations, gcmr_workplaces,
                                          gcmr_residential) %>% 
  mutate(lag_confirmed = lag(confirmed, 1),
         lag_deaths = lag(deaths, 1),
         daily_confirmed = case_when(date == "2020-02-24" ~ confirmed,
                                     date > "2020-02-24" ~ confirmed - lag_confirmed),
         daily_deaths = case_when(date == "2020-02-24" ~ deaths,
                                     date > "2020-02-24" ~ deaths - lag_deaths),
         infection_rate = daily_confirmed / population * 1000000,
         daily_confirmed_proj = daily_confirmed*10,
         confirmed_est = confirmed*10,
         daily_deaths_proj = daily_deaths*2,
         deaths_est = deaths*2,
         infection_rate_proj = daily_confirmed_proj/ population * 1000000)
```


```{r budge-data}
budget <- read_csv("data/budget.csv") 


budget <- budget %>% mutate(index = gsub('\\D+','', budget$`COFOG based budget`),
         budget = gsub('\\d+', '', budget$`COFOG based budget`))

budget_new <- budget %>%
  filter(index %in% c(1:9)) %>%
  select(-`COFOG based budget`, -index) %>%
  rename(Budget = budget,
         "2019" = "1399",
         "2020" = "1400",
         "2021" = "1401",
         "2022" = "1402")

col_order <- c("Budget", "2019", "2020", "2021", "2022")

budget_table<-  budget_new[, col_order]
```


```{r health-data}

health_profile <- read_csv("data/health_profile.csv")

health_facility <- read_csv("data/health_care.csv")%>%
  filter(Indicators %in% c("Total Number of Hospitals",
                           "Bed in hospitals (government)",
                           "Bed  per ten thousand person",
                           "Doctors per ten thousand person"))

health <- rbind(health_profile, health_facility) %>%
  filter(Indicators != "Male",
         Indicators != "Female")

```

<br>


**Analysis Tools**

The analysis is performed using R [@r]. The data is wrangled mainly using the tidyverse function [@tidyverse]. Further, the data is visualized using the function in `ggplot2` [@ggplot2], `gghighlight` [@gghighlight] and `ggrepel` [@ggrepel]. I also used `gt` [@gt] and and `scales` [@scales].


## COVID-19 Cases in Afghanistan


To analyze the COVID-19 situation in Afghanistan, I used the cumulative numbers of confirmed cases and deaths. It is because I only want to see the general condition of COVID-19 in this country.


```{r covid-plot-dat}

covid_plot_dat <- afghan_data_new %>%
  select(date, confirmed, deaths,
         daily_confirmed, daily_deaths, daily_confirmed_proj, daily_deaths_proj,
         confirmed_est, deaths_est) %>%
  rename(`confirmed` = confirmed,
         `deaths` = deaths,
         `estimated cases` = confirmed_est,
         `estimated deaths` = deaths_est) %>%
  select(date, `confirmed`,
         `deaths` ,
         `estimated cases`,
         `estimated deaths`)

covid_plot_dat<- covid_plot_dat %>%
  pivot_longer(c(2:5),
               names_to = "indicator",
               values_to = "number")

```


<center>

```{r covid-plot, fig.cap= "COVID-19 confirmed cases and deaths in Afghanistan 24 February to 15 September 2020"}

cols1 <- c("confirmed" = "#FF6961", "deaths" = "#875C36")

data_ends <- filter(covid_plot_dat, indicator %in% c("confirmed", "deaths"),
                                                     date == "2020-09-15")

covid_plot <- ggplot(filter(covid_plot_dat, indicator %in% c("confirmed", "deaths")), aes(x=date, y=number, group=indicator)) + 
  geom_line(aes(linetype=indicator, color=indicator)) +
  theme(legend.position = "bottom",
        legend.justification='left',
        legend.direction='horizontal') +
  scale_y_continuous(breaks = seq(0, 40000, by=10000)) +
  theme(panel.grid.major = element_line(size = 0.5, linetype = 'solid',
                                colour = "#D7DBDD"), 
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "grey92"),
        plot.title.position = "plot",
        plot.title = element_text(size = 12),
        plot.margin = margin(10, 10, 10, 10),
        legend.background = element_rect(fill = "grey92"),
        legend.title = element_blank()) +
  scale_colour_manual(values = cols1) +
  ggtitle("Cumulative Confirmed Cases and Deaths of Covid-19 in Afghanistan") +
  xlab("Date") +
  ylab("Cumulative cases in thousand") +
  labs(caption = "Source: JHU CSSE") +
  geom_text_repel(
    aes(label = number), data = data_ends,
    fontface ="plain", color = "black", size = 3
    )


#cols2 <- c("deaths" = "#875C36", "estimated deaths" = "#E1B894")
#p2 <- ggplot(filter(covid_plot_dat, indicator %in% c("deaths", "estimated deaths")), aes(x=date, y=number, group=indicator)) +
  #geom_line(aes(linetype=indicator, color=indicator)) +
  #theme(legend.position = "bottom",
        #legend.justification='left',
        #legend.direction='horizontal',
        #panel.grid.major = element_line(size = 0.5, linetype = 'solid',
                                #colour = "#D7DBDD"), 
        #panel.grid.minor = element_blank(),
        #plot.background = element_rect(fill = "grey92"),
        #plot.title.position = "plot",
        #plot.title = element_text(size = 12),
        #plot.margin = margin(10, 10, 10, 10),
        #legend.background = element_rect(fill = "grey92"),
        #legend.title = element_blank()) +
  #scale_colour_manual(values = cols2) +
  #ggtitle("B)") +
  #xlab("Date") +
  #ylab("Cumulative deaths") +
  #labs(caption = "Source: JHU CSSE")


#p1 + p2 + plot_layout(ncol = 2) 

covid_plot

  
```
</center>




```{r}
last_data_conf <- filter(covid_plot_dat, indicator == "confirmed" & date == "2020-09-15")
last_conf <- last_data_conf$number

last_data_deaths <- filter(covid_plot_dat, indicator == "deaths" & date == "2020-09-15")
last_deaths <- last_data_deaths$number

infection_rate <- round(last_conf / afghan_data_new$population[1] * 1000000, 2)
fatality_rate <- round(last_deaths/last_conf*100, 2)
```


Figure \@ref(fig:covid-plot) shows that until 15 September 2020, the number of COVID-19 cases in Afghanistan has reached `r comma(last_conf)`, or equivalent to `r infection_rate` cases per one million population. We can also see that the curve flattened. 

Regarding the death case, as of 15 September 2020, the number is `r comma(last_deaths)` with `r fatality_rate` percent of case fatality rate. However, we could not conclude anything from this indicator because the WHO said that the fatality rate varies in each country from 0.1% to more than 25%. The calculation method is also still rough [@roughdata]. 

If we only look at the confirmed case number, we might conclude that Afghanistan has handled this pandemic well, given the evidence that the curve started to flatten. However, how reliable is the data? Of course, we cannot see this indicator alone. We also have to take other indicators into account, namely the testing rate and the positive rate. Unfortunately, it is tough to find Afghanistan's data for these indicators. [Our World In Data](https://ourworldindata.org/coronavirus-testing), the source of testing data in `tidycovid19` package also does not have this indicator data for Afghanistan.

Further, according to the WHO, the actual rate of transmission is mostly underestimated since the disease is sometimes asymptomatic [@roughdata]. The low testing capacity in the country then exacerbated the situation. According to Nicholas Bishop, an expert of the United Nations International Organization for Migration (IOM), as per June 2020, Afghanistan only held 646 tests per million population, bringing it to be one of the countries with lowest testing in the world [@bishop]. He also added that the true cases of COVID-19 in Afghanistan could be millions since there is a significant level of community transmissions across 34 provinces. Similarly, a survey conducted by KabulNow concluded that the actual cases in Kabul could be ten times more than the official record [@kabulnow]. 


## Why it could be worse?

Assuming that the actual cases are much higher than the official record, we could see an already bad situation in this war-torn country. Unfortunately, as I mentioned in the introduction, Afghan people are already miserable enough because of the conflict. Hence in this part, I would emphasize the conflict effects that made the pandemic worse in this country. 


**Health Care Vulnerability**

The prolonged conflict in the past 18 years in Afghanistan has shifted the government focus to defense and safety affairs. Figure \@ref(fig:budget-plot) suggests that more than 40 percent of the government’s budget plan in 2020 is for defense and public order and safety. The fund for public health only accounts for 4.4 percent, much lower than the other budget.

```{r budget-plot-dat}
budget_plot_dat <- budget_table %>%
  select(Budget, `2020`) %>%
  mutate(percent_budget = round(`2020`/sum(`2020`)*100, 2))
```

<center>
```{r budget-plot, fig.cap = "Afghanistan's plan of budget distribution in 2020"}

budget_plot <- ggplot(budget_plot_dat) +
  geom_col(aes(x= percent_budget, y = reorder(Budget, percent_budget)), fill = "#009bda") +
  ylab("") +
  xlab("") +
  geom_text(aes(x= percent_budget, y= reorder(Budget, percent_budget), label = percent_budget, hjust = -0.1)) +
  theme_void()+
  theme(axis.text.y = element_text(size = 8.5, color = "black", hjust = 1),
        plot.title.position = "plot",
        plot.title = element_text(size = 12)) +
  ggtitle("Afghanistan Budget Plan for 2020 (%)") +
  labs(caption = "Source: Afghansitan Ministry of Finance") +
  gghighlight(percent_budget == 4.4)

budget_plot

```
</center>

<details><summary> Expand here if you want to see the detailed number of the budget </summary>

```{r}

gt(budget_table) %>%
  tab_header(title = md("Afghanistan Budget Outlook (in Afghan Afghani)")) %>%
  fmt_number(columns = starts_with("20"), suffixing = TRUE) %>%
  tab_footnote(
    footnote = "Budget plan",
    locations = cells_column_labels(c("2020", "2021", "2022")))
  

```

</details>

<br>

```{r health-table}

health_table <- health %>%
  filter(Indicators %in% c("Number of  doctors",
                           "Number of Health Associate Professional",
                           "Number of  beds in public hospitals",
                           "Total Number of Hospitals",
                           "Bed  per ten thousand person",
                           "Doctors per ten thousand person")) 

gt(health_table) %>%
  tab_header(title = md("Health Care Profile of Afghanistan 2017-2019")) %>%
  tab_source_note(source_note = md("Source: Afghanistan Statistical Year Book 2019, NSIA, page 131-132")) %>%
  tab_style(style = list(cell_fill(color = "#a0d0de")),
    locations = cells_body(
      columns = vars(`2019`),
      rows = `2019` <= 5)) 

```


Further the health care facility in Afghanistan could be considered as inappropriate. In the three years of 2017 to 2019, there is no significant progress in health facilities provision in Afghanistan (refer to Table 1). The number of health associate professional even decreased in 2019. Further, the latest available data show that there are only three doctors for every 10,000 Afghans. Meanwhile, there are only five hospital beds per 10,000 population. 

This reality is even more bitter because health facilities and their workers are often the targets of non-state armed groups. According to the UN Assistance Mission in Afghanistan report, 12 deliberate attacks on health care have happened during the pandemic [@DeviSharmila2020Athc]. One of them was carried out by armed men who attacked Dasht-e-Barchi hospital in Kabul on 12 May 2020. This attack killed 24 people, including 16 women and two babies. After this incident, Médecins Sans Frontières (MSF), a medical humanitarian organization, stopped its activities at this hospital, which means that one health facility was depleted, even though the need is increasing due to the pandemic.


<br>

**Economic Vulnerability**


Conflict does affect not only health care but also the economy. Afghanistan is one of the world's poorest countries, and the perpetuated civil war is the culprit [@wikipoverty] and [@xinhuanet].  Even before COVID-19 strikes, Afghans lived with an average income of less than 2 US dollars a day [@DeviSharmila2020Athc]. With this poverty level, the government will find it challenging to continue to apply the restriction that has been going on since early March [@worldbankpov]. If this happened, there would be a new threat, a hunger [@guardian]. 


```{r mob-data}
mobility <- afghan_data_new %>%
  select(date,
         gcmr_grocery_pharmacy,
         gcmr_retail_recreation,
         gcmr_parks,
         gcmr_residential,
         gcmr_transit_stations,
         gcmr_workplaces) %>%
  rename(`grocery and pharmacy` = gcmr_grocery_pharmacy,
         `retail and recreation` = gcmr_retail_recreation,
         `parks` = gcmr_parks,
         `residential` = gcmr_residential,
         `transit station` = gcmr_transit_stations,
         `workplaces` = gcmr_workplaces) %>%
  pivot_longer(c(2:7), names_to = "gmcr", values_to = "index")
```

```{r int-plot, eval = FALSE}

# declare `city` as the SQL 'query by' column
tx <- highlight_key(mobility, ~gmcr)

# initiate a plotly object
base <- plot_ly(tx, color = I("black")) %>% 
  group_by(gmcr)

# create a time series of median house price
base %>%
  group_by(gmcr) %>%
  add_lines(x = ~date, y = ~index) %>%
  highlight(
  on = "plotly_click",
  selectize = TRUE, 
  dynamic = TRUE, 
  persistent = TRUE
)
```

<center>
```{r mob-data-plot, fig.cap= "Community mobility index in Afghanistan amid COVID-19"}

ggplot(transform(mobility,
      gmcr=factor(gmcr,levels=c("grocery and pharmacy",
                                "retail and recreation",
                                "parks",
                                "residential",
                                "transit station",
                                "workplaces")))) +
  geom_line(aes(x = date,
                y = index,
                colour = gmcr)) +
  facet_wrap(~gmcr) +
  theme(legend.position = "none",
        panel.grid.major = element_line(size = 0.5, linetype = 'solid',
                                colour = "#D7DBDD"), 
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "grey92"),
        plot.title.position = "plot",
        plot.title = element_text(size = 12),
        plot.margin = margin(10, 10, 10, 10)) +
  labs(x = "Date",
       y = "Google Mobility Index",
       caption = "Source: Google Community Mobility Report") +
  ggtitle("Google Mobility Index by Place in Afghanistan")
```
</center>

We can infer from Figure \@ref(fig:mob-data-plot) that when the restriction was implemented, there was a decrease in community mobility almost in every place. However, in early July, the mobility to grocery and pharmacy, as well as retails and recreation, has increased again. It is because most of the population works as shopkeepers, street or retail sales, or daily labor in construction, agriculture, or personal services.

The poverty as the result of conflict made Afghans have no choice but to survive. That being said, returning to work amid a pandemic will undoubtedly increase the risk of being infected by COVID-19.

## Conclusion

## Acknowledgments {.appendix}

<span>The preview photo by <a href="https://unsplash.com/@kingmaphotos?utm_source=unsplash&amp;utm_medium=referral&amp;utm_content=creditCopyText">Jeff Kingma</a> on <a href="https://unsplash.com/?utm_source=unsplash&amp;utm_medium=referral&amp;utm_content=creditCopyText">Unsplash</a></span>

- Distill

- Google Translate









