---
title: "Covid and Conflict: Why the pandemic could be worse in Afghanistan?"
description: |
  This article aims to convey the reason why the conflicted country, in this case is Afghanistan, might be hit hard by Covid-19.
author:
  - name: Dewi Lestari Amaliah
    url: {}
date: 09-15-2020
output:
  distill::distill_article:
    self_contained: false
    toc: true
bibliography: references2.bib
categories:
  - COVID-19
  - Data Analysis
draft: TRUE
preview: war.jpg
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(tidycovid19)
library(plotly)
library(hrbrthemes)
library(waffle)
library(gt)
library(DT)
library(kableExtra)
library(gghighlight)
library(crosstalk)
library(ggrepel)
```


```{css, echo = FALSE}

html, body, p {
  
  font-family: 'Lato', sans-serif;
  font-weight: 200;
  line-height: 1.3; 
  font-size: 1.0em;
  color: black;
  font-style: normal;
  
}

table.dataTable tbody td {
    padding: 12px 10px;
}


d-article table th, d-article table td {
    font-size: 13px;
    padding: 12px 10px;
}

```



## Introduction

Since being declared as a global pandemic by the World Health Organization (WHO) on 12 March 2020, Covid-19 has affected 213 countries, including the conflict-torn country, Afghanistan [@countries]. This country has its first confirmed case recorded on 24 February 2020 in Herat after a citizen who had just returned from Iran contracted with the virus [@firstcase].

Viruses are indeed indiscriminate. Unfortunately, the presence of Covid-19 amid this war-shattered country seems to rub salt into the wound. According to the United Nations Office for the Coordination of Humanitarian Affairs (OCHA) data, as of 6 September 2020, there were already 151,190 people displaced by the conflict [@ocha]. Further, in 2019, civilian casualties due to the conflict has surpassed 100,000 [@vulnerable]. The Secretary-General’s Special Representative for Afghanistan and head of the UN Assistance Mission in Afghanistan (UNAMA) said as follows:  

> "Almost no civilian in Afghanistan has escaped being personally affected in some way by the ongoing violence" - *Tadamichi Yamamoto* [@vulnerable] 

In other words, the Afghan people are already vulnerable before Covid-19 attacks. 

This article aims to convey the reasons why the conflict made the country might be hit hard by Covid-19. The analysis would be focus on two aspects, namely the health care and the population vulnerability.  


## Data Source

In doing analysis, I used several data sets from various sources. For the Covid-19 situation in Afghanistan, I used the data of confirmed cases from John Hopkins University Center for System Science and Engineering (JHU CSSE) [@jhu] and the population data from the World Bank [@population]. I also used the mobility data from Google Covid-19 Community Mobile Report [@googlecmr] to analyze the population vulnerability. These data have been compiled in the `tidycovid19` R package by Joachim Gassen [@tidycovi19], hence I directly downloaded it from this package. For the health care vulnerability analysis, I mainly used the data of health sector profile from the Afghanistan National Statistics and Information Authority [@nsia] and the government national budget data from the Afghanistan Ministry of Finance [@mof].  


```{r covid-data}
# download data for all countries available in the package and all of the compiled data
merged_data <- download_merged_data(cached = TRUE, silent = TRUE)

# filter the data only for Afghanistan 
afghan_data <- merged_data %>% filter(iso3c == "AFG") %>%
  select(-ecdc_cases, -ecdc_deaths) %>%
  filter(confirmed != 0)
```

```{r afghan-data}
afghan_data_new <- afghan_data %>% select(iso3c, country, date, deaths, confirmed, population, gcmr_retail_recreation,
                                          gcmr_grocery_pharmacy, gcmr_parks, gcmr_transit_stations, gcmr_workplaces,
                                          gcmr_residential) %>% 
  mutate(lag_confirmed = lag(confirmed, 1),
         lag_deaths = lag(deaths, 1),
         daily_confirmed = case_when(date == "2020-02-24" ~ confirmed,
                                     date > "2020-02-24" ~ confirmed - lag_confirmed),
         daily_deaths = case_when(date == "2020-02-24" ~ deaths,
                                     date > "2020-02-24" ~ deaths - lag_deaths),
         infection_rate = daily_confirmed / population * 1000000,
         daily_confirmed_proj = daily_confirmed*10,
         confirmed_est = confirmed*10,
         daily_deaths_proj = daily_deaths*2,
         deaths_est = deaths*2,
         infection_rate_proj = daily_confirmed_proj/ population * 1000000)
```


```{r budge-data}
budget <- read_csv("data/budget.csv") 


budget <- budget %>% mutate(index = gsub('\\D+','', budget$`COFOG based budget`),
         budget = gsub('\\d+', '', budget$`COFOG based budget`))

budget_new <- budget %>%
  filter(index %in% c(1:9)) %>%
  select(-`COFOG based budget`, -index) %>%
  rename(Budget = budget,
         "2019" = "1399",
         "2020" = "1400",
         "2021" = "1401",
         "2022" = "1402")

col_order <- c("Budget", "2019", "2020", "2021", "2022")

budget_table<-  budget_new[, col_order]
```


```{r health-data}

health_profile <- read_csv("data/health_profile.csv")

health_facility <- read_csv("data/health_care.csv")%>%
  filter(Indicators %in% c("Total Number of Hospitals",
                           "Bed in hospitals (government)",
                           "Bed  per ten thousand person",
                           "Doctors per ten thousand person"))

health <- rbind(health_profile, health_facility) %>%
  filter(Indicators != "Male",
         Indicators != "Female")

```

<br>


**Analysis Tools**

The analysis is performed using R [@r]. The data is wrangled mainly using the tidyverse function [@tidyverse]. Further, the data is visualized using the function in `ggplot2` [@ggplot2], plotly [@plotly] and `waffle` [@waffle].


## Covid-19 Cases in Afghanistan

The confirmed cases data from `tidycovid19` is provided in cumulative-wise, hence I wrangled the data first to get the number of daily cases. 


```{r covid-plot-dat}

covid_plot_dat <- afghan_data_new %>%
  select(date, confirmed, deaths,
         daily_confirmed, daily_deaths, daily_confirmed_proj, daily_deaths_proj,
         confirmed_est, deaths_est) %>%
  rename(`confirmed` = confirmed,
         `deaths` = deaths,
         `estimated cases` = confirmed_est,
         `estimated deaths` = deaths_est) %>%
  select(date, `confirmed`,
         `deaths` ,
         `estimated cases`,
         `estimated deaths`)

covid_plot_dat<- covid_plot_dat %>%
  pivot_longer(c(2:5),
               names_to = "indicator",
               values_to = "number")

```


```{r covid-plot}

cols1 <- c("confirmed" = "#FF6961", "deaths" = "#875C36")

data_ends <- filter(covid_plot_dat, indicator %in% c("confirmed", "deaths"),
                                                     date == "2020-09-15")

ggplot(filter(covid_plot_dat, indicator %in% c("confirmed", "deaths")), aes(x=date, y=number, group=indicator)) + 
  geom_line(aes(linetype=indicator, color=indicator)) +
  theme(legend.position = "bottom",
        legend.justification='left',
        legend.direction='horizontal') +
  scale_y_continuous(breaks = seq(0, 40000, by=5000)) +
  theme(panel.grid.major = element_line(size = 0.5, linetype = 'solid',
                                colour = "#D7DBDD"), 
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "grey92"),
        plot.title.position = "plot",
        plot.title = element_text(size = 12),
        plot.margin = margin(10, 10, 10, 10),
        legend.background = element_rect(fill = "grey92"),
        legend.title = element_blank()) +
  scale_colour_manual(values = cols1) +
  ggtitle("Cumulative Confirmed Cases and Deaths of Covid-19 in Afghanistan") +
  xlab("Date") +
  ylab("Cumulative cases in thousand") +
  labs(caption = "Source: JHU CSSE") +
  geom_text_repel(
    aes(label = number), data = data_ends,
    fontface ="plain", color = "black", size = 3
    )




#cols2 <- c("deaths" = "#875C36", "estimated deaths" = "#E1B894")
#p2 <- ggplot(filter(covid_plot_dat, indicator %in% c("deaths", "estimated deaths")), aes(x=date, y=number, group=indicator)) +
  #geom_line(aes(linetype=indicator, color=indicator)) +
  #theme(legend.position = "bottom",
        #legend.justification='left',
        #legend.direction='horizontal',
        #panel.grid.major = element_line(size = 0.5, linetype = 'solid',
                                #colour = "#D7DBDD"), 
        #panel.grid.minor = element_blank(),
        #plot.background = element_rect(fill = "grey92"),
        #plot.title.position = "plot",
        #plot.title = element_text(size = 12),
        #plot.margin = margin(10, 10, 10, 10),
        #legend.background = element_rect(fill = "grey92"),
        #legend.title = element_blank()) +
  #scale_colour_manual(values = cols2) +
  #ggtitle("B)") +
  #xlab("Date") +
  #ylab("Cumulative deaths") +
  #labs(caption = "Source: JHU CSSE")


#p1 + p2 + plot_layout(ncol = 2) 



  
```




## Why it could be worse?

**Health Care Vulnerability**


```{r budget-plot-dat}
budget_plot_dat <- budget_table %>%
  select(Budget, `2020`) %>%
  mutate(percent_budget = round(`2020`/sum(`2020`)*100, 2))
```

```{r}
budget_plot <- ggplot(budget_plot_dat) +
  geom_col(aes(x= percent_budget, y = reorder(Budget, percent_budget)), fill = "#009bda") +
  ylab("") +
  xlab("") +
  geom_text(aes(x= percent_budget, y= reorder(Budget, percent_budget), label = percent_budget, hjust = -0.1)) +
  theme_void()+
  theme(axis.text.y = element_text(size = 8.5, color = "black", hjust = 1),
        plot.title.position = "plot",
        plot.title = element_text(size = 12)) +
  ggtitle("Afghanistan Budget Plan for 2020 (%)") +
  labs(caption = "Source: Afghansitan Ministry of Finance") +
  gghighlight(percent_budget == 4.4)

budget_plot

```

<details><summary> Expand here if you want to see the detailed number of the budget </summary>

```{r}

gt(budget_table) %>%
  tab_header(title = md("Afghanistan Budget Outlook (in Afghan Afghani)")) %>%
  fmt_number(columns = starts_with("20"), suffixing = TRUE) %>%
  tab_footnote(
    footnote = "Budget plan",
    locations = cells_column_labels(c("2020", "2021", "2022")))
  

```

</details>

<br>

```{r}

health_table <- health %>%
  filter(Indicators %in% c("Number of  doctors",
                           "Number of Health Associate Professional",
                           "Number of  beds in public hospitals",
                           "Total Number of Hospitals",
                           "Bed  per ten thousand person",
                           "Doctors per ten thousand person")) 

gt(health_table) %>%
  tab_header(title = md("Health Care Profile of Afghanistan 2017-2019")) %>%
  tab_source_note(source_note = md("Source: Afghanistan Statistical Year Book 2019, NSIA, page 131-132")) %>%
  tab_style(style = list(cell_fill(color = "#a0d0de")),
    locations = cells_body(
      columns = vars(`2019`),
      rows = `2019` <= 5)) 

```









**Population Vulnerability**


```{r}
mobility <- afghan_data_new %>%
  select(date,
         gcmr_grocery_pharmacy,
         gcmr_retail_recreation,
         gcmr_parks,
         gcmr_residential,
         gcmr_transit_stations,
         gcmr_workplaces) %>%
  rename(`grocery and pharmacy` = gcmr_grocery_pharmacy,
         `retail and recreation` = gcmr_retail_recreation,
         `parks` = gcmr_parks,
         `residential` = gcmr_residential,
         `transit station` = gcmr_transit_stations,
         `workplaces` = gcmr_workplaces) %>%
  pivot_longer(c(2:7), names_to = "gmcr", values_to = "index")
```



```{r int-plot, eval = FALSE}

# declare `city` as the SQL 'query by' column
tx <- highlight_key(mobility, ~gmcr)

# initiate a plotly object
base <- plot_ly(tx, color = I("black")) %>% 
  group_by(gmcr)

# create a time series of median house price
base %>%
  group_by(gmcr) %>%
  add_lines(x = ~date, y = ~index) %>%
  highlight(
  on = "plotly_click",
  selectize = TRUE, 
  dynamic = TRUE, 
  persistent = TRUE
)
```


```{r}
ggplot(transform(mobility,
      gmcr=factor(gmcr,levels=c("grocery and pharmacy",
                                "retail and recreation",
                                "parks",
                                "residential",
                                "transit station",
                                "workplaces")))) +
  geom_line(aes(x = date,
                y = index,
                colour = gmcr)) +
  facet_wrap(~gmcr) +
  theme(legend.position = "none",
        panel.grid.major = element_line(size = 0.5, linetype = 'solid',
                                colour = "#D7DBDD"), 
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "grey92"),
        plot.title.position = "plot",
        plot.title = element_text(size = 12),
        plot.margin = margin(10, 10, 10, 10)) +
  labs(x = "Date",
       y = "Google Mobility Index",
       caption = "Source: Google Community Mobility Report") +
  ggtitle("Google Mobility Index by Place in Afghanistan")
```


## Conclusion

## Acknowledgments {.appendix}

<span>The preview photo by <a href="https://unsplash.com/@kingmaphotos?utm_source=unsplash&amp;utm_medium=referral&amp;utm_content=creditCopyText">Jeff Kingma</a> on <a href="https://unsplash.com/?utm_source=unsplash&amp;utm_medium=referral&amp;utm_content=creditCopyText">Unsplash</a></span>







